
\section{DTrace Instruction Version}

This specification describes the DTrace Intermediate Format version 2, as
shipped in Solaris \textbf{XXXRW}, FreeBSD \textbf{XXXRW}, and Mac OS X
\textbf{XXXRW}.

\section{The DIF Interpreter}

The DTrace Intermediate Format (DIF) interpreter executes instructions
on behalf of D scripts that are associated with predicates and
actions.  DIF is a simple, RISC-like, instruction set where each
instruction consists of a 32-bit, native-endian integer whose most
significant 8 bits contain an opcode allowing the remainder of the
instruction to be decoded.  Interpretation is executed in a loop
within the \subroutine{dtrace_dif_emulate} subroutine, which has, as
its first argument, a pointer to a DIF Object or
\struct{dtrace_difo_t}.  Instructions are executed one at a time,
until they are exhausted or an error causes interpretation to end. The DIF
objects are verified in the \subroutine{dtrace_difo_validate} subroutine and
the DIF interpreter ignores any bounds checking within the
\subroutine{dtrace_dif_emulate} subroutine precisely because
\subroutine{dtrace_difo_validate} performs the necessary checks.

\subsection{Registers}
\label{sec:dif-registers}

The DTrace virtual machine is made up of eight (8) integer registers
and eight (8) tuple registers.  The 0th integer register is always contains
the value zero (0).  All operations are carried out using registers
\registerop{r1} and \registerop{r2} as operands and \registerop{rd} as
the destination for all results.

\subsection{Math Instructions}
\label{sec:dif-math}

Instructions for mathematical operations in DIF have no concept of
over or underflow.  The division instructions set a flag to indicate a
division by zero error.

\subsection{Comparison and Test Instructions}
\label{sec:dif-cmp-tst}

\begin{table}
  \centering
    \begin{tabular}{|l|l|}
      \hline
      Variable & Meaning\\
      \hline
      cc\_r & Value of $r1 - r2$\\
      cc\_n & Comparison result is negative. \\
      cc\_z & Comparison result is 0.\\
      cc\_v & Overflow occurred.\\
      cc\_c & Is $r1 < r2$?\\
      \hline
  \end{tabular}
\label{tbl:cmp-vars}
\caption{Mathematical Operation Result Bits}
\end{table}

DIF has three instructions, \instruction{cmp}, \instruction{scmp} and
\instruction{tst} which can set various result flags, shown in
Table~\ref{tbl:cmp-vars}, which are later used by the branch
instructions.  The result flags are never returned directly to the
calling DIF program but are only used internally by the interpretation
routine.

\textbf{XXXRW: Something about instructions}

\subsection{Branching Instructions}

DIF has eleven split into two types: signed and unsigned branching instructions.
The signed branching instructions take into account that the number may be
negative, while the unsigned instructions are meant to be used with exclusively
positive numbers. One thing all of the branching instructions have in common is
that they load the new \register{pc} register from the \registerop{label} field
in the Branch Format described in Subsection~\ref{subsec:b-format}.

\subsection{Subroutine Calls}
\label{sec:subroutines}

DTrace provides an extensive set of subroutines for use by D
programs.  The subroutines are implemented within the kernel code via
a set of functions which are centrally dispatched from the
\function{dtrace_dif_subr} function.  Within DIF subroutines are
triggered via the \instruction{CALL} instruction.  Any subroutine that
is provided to DTrace \em{must} go via these mechanisms.

\begin{figure}
\begin{quote}
  Notice that we don't bother validating the proper number of
  arguments or their types in the tuple stack.  This isn't needed
  because all argument interpretation is safe because of our load
  safety -- the worst that can happen is that a bogus program can
  obtain bogus results.
\end{quote}
\label{fig:argcheck}
\end{figure}

According to a comment in the code, see Figure~\ref{fig:argcheck},
argument checks are not carried out when a subroutine is called.


\textbf{XXXRW: Something about returning}

\textbf{XXXRW: Something about DIF registers, \nregs{}}

\textbf{XXXRW: Something about the zero register}

\textbf{XXXRW: Something about implied status bits}

\textbf{XXXRW: Something about scratch memory (alloca, user copyin)}

\textbf{XXXRW: Something about the tuple stack}

\textbf{XXXRW: Something about kernel memory access}

\textbf{XXXRW: Something about user memory access}

\textbf{XXXRW: Something about ``by reference''}

\textbf{XXXRW: Something about the integer table}

\textbf{XXXRW: Something about the string table}

\subsection{Local Variables}
\label{sec:local-vars}

Local variables are local to the D program clause.  In a D program
these variables are referenced using the \verb|this->| syntax.  

\subsection{Thread Local Variables}
\label{sec:thread-local-vars}

Thread local variables are usable in multiple D program clauses.  In a
D program the thread local variables are referenced using the
\verb|self->| syntax.

\subsection{Global Variables}
\label{sec:globals-vars}

Global variables are global to all the clauses in a D script and are references
to simple names within the D script.  Space for global variables is statically
allocated on each invocation of a script. Additionally, global variables are
identified using the modified register format as describ ed in
Subsection~\ref{subsec:r-format} the case of arrays and the wide-immediate
format which will be described in Subsection~\ref{subsec:w-format} in the case
of scalar variables inside of the \subroutine{dtrace_dif_variable} subroutine.

\textbf{XXXRW: Something about scalars}

\textbf{XXXRW: Something about arrays}

\textbf{XXXRW: Something about aggregations}

\textbf{XXXRW: Something about exceptions}

\section{Instruction Format}

Each instruction consists of a 32-bit, native-endian integer whose most
significant 8 bits contain an opcode allowing the remainder of the instruction
to be decoded.
To ease parsing, three major formats (R, B, and W) are used for all DTrace
instructions, capturing different types of operations: register-to-register
instructions accepting zero or more register operands; branch instructions
accepting a target label as a single operand; and wide-immediate instructions
that accept a 16-bit immediate used to capture both small constant values and
also indices into various tables.

\subsection{Register Format (R-Format)}
\label{subsec:r-format}
This format accepts zero or more register operands, supporting instructions
that include arithmetic and boolean operations, comparison and test
operations, load and store operations, tuple-stack operations, and the no-op
instruction.

\begin{center}
\begin{bytefield}[endianness=big,bitformatting=\scriptsize]{32}
\bitheader{0,7,8,15,16,23,24,31}\\
\bitbox{8}{op}
\bitbox{8}{r1}
\bitbox{8}{r2}
\bitbox{8}{rd}
\end{bytefield}
\end{center}

\begin{description}
\item[op] Mandatory 8-bit operation identifier
\item[r1, r2] Optional source registers providing input values to the
  operation
\item[rd] Optional destination register acting as the destination of the
  operation
\end{description}


A modified version of the Register Format is used when loading and
storing data in array variables in DTrace. The main difference between the
regular Register Format and the modified one used for arrays, is that the
\registerop{r1} register location is used as the variable identificator, and the
\registerop{r2} register location is used as an optional index in the array.

\begin{center}
\begin{bytefield}[endianness=big,bitformatting=\scriptsize]{32}
\bitheader{0,7,8,15,16,23,24,31}\\
\bitbox{8}{op}
\bitbox{8}{var}
\bitbox{8}{ndx}
\bitbox{8}{rd}
\end{bytefield}
\end{center}

\begin{description}
	\item[op] Mandatory 8-bit operation identifier
	\item[var] The variable identifier
	\item[ndx] Optional index of the array variable to be accessed
	\item[rd] Optional destination register acting as the destination of the
operation
\end{description}

\subsection{Branch Format (B-Format)}
\label{subsec:b-format}

This format accepts a single 24-bit integer operand identifying the label that
is the branch target.
It is used solely for the \textbf{BRANCH} instruction.

\begin{center}
\begin{bytefield}[endianness=big,bitformatting=\scriptsize]{32}
\bitheader{0,23,24,31}\\
\bitbox{8}{op}
\bitbox{24}{label}
\end{bytefield}
\end{center}

\begin{description}
\item[op] Mandatory 8-bit operation identifier
\item[label] Mandatory 24-bit integer label
\end{description}

\subsection{Wide-Immediate Format (W-Format)}
\label{subsec:w-format}

This format accepts an 8-bit register and 16-bit integer argument (frequently
an index).
It is used for a range of instructions including those to load values from
integer and string constant tables, as well as those that store scalar values in
variables.

\begin{center}
\begin{bytefield}[endianness=big,bitformatting=\scriptsize]{32}
\bitheader{0,7,8,23,24,31}\\
\bitbox{8}{op}
\bitbox{16}{idx}
\bitbox{8}{rs\textbar rd}
\end{bytefield}
\end{center}

\begin{description}
\item[op] Mandatory 8-bit operation identifier
\item[idx] Mandatory 16-bit integer index
\item[rs\textbar rd] Optional 8-bit register acting as the source or
  destination of the operation
\end{description}

\section{Instructions}

Chapter~\ref{chap:dtrace-instruction-reference} contains a comprehensive
description of DTrace's instructions.

\section{Subroutines}

Chapter~\ref{chap:dtrace-subroutines} contains a comprehensive description of
DTrace's built-in subroutines.
